Formal Verification of Real Models from Platforms Using The B-method

### 2009-04-22 papers/FM2009/source/paper.tex

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

%\usepackage{float}

@@

\begin{abstract}
This paper describes an approach to model the microcontroller platforms. More specifically, it shows details about
the Z80 model. The model is created using the B method; it applies math and logic concepts to describe the
characteristics from platforms. Therefore, this model can be used in platforms projects to document, build
simulators, verify properties about the model and verify the software in assembly level. The verification in
assembly is utility more important, because this allow the developing software faithful to the algorithm language.
Finally, this paper show some relevant techniques used in models and to describe several details.
\end{abstract}
%
\section{Introduction about hardware models in B}
[One page]
This paper show a model in B from a microcontrooller.
[EXPLAIN THE MOTIVATION]
 [When a error is found at project of plataform the PREJUIZOS are big ]
 [ In critical system, when a error apper in software it can make several PREJUIZOS  ]

[[Other microcontrollers are in building proccess: 8051, Arm..]]
[Some smal details of the model are omitted in order to simplify the explanation.]
[Smal intrudctio about B]
[Structure from paper]

%

@@

\subsubsection{ Desing components B }
[half page]
[Explicar como as coias são modeladas]

@@

The main module includes a instance of memory module and accesses the definitions from: basic data types modules,
$\mathit{ALU}$ and $\mathit{Power2}$ \footnote{The $\mathit{Power2}$ module has the basics definitions to help the
theorems prover about calculus of power.}.

@@

There are many functions defined in the module ALU. In general, these functions take basic definitions to build
new specific functions. The function half8UCHAR is used to get the half part of UCHAR value. It is importante to
know the half carry and it is used in the $\mathit{add8UCHAR}$.

@@

Every logic functions defined in the $\mathit{BYTE}$ and $\mathit{BV16}$ module are included in the $\mathit{ALU}$
module. Then, it was not necessary rewrite these functions.

\subsection{Interrupts}

% Visão Geral
In the Z80, the interrupts allows that the devices suspend a routine from CPU and start another service routine.
This service routine can exchange data or signals between CPU and external devices. When a routine is finished, then
the CPU come back to the last routine that was interrupted.

@@

% $updateStack(\{ (sp\_minus\_two \fun low\_pc ),(sp\_minus\_one
%    \fun high\_pc ) \}) ||\\  sp:= sp\_minus\_two || pc := 66H || iff1:=0 || 
%    iff2:= iff1 $
\begin{sloppypar}
\bf updateStack\rm (\rm \{ \rm (\it sp\_minus\_two  $\mapsto$  \it pc\_low\rm )\rm ,\rm (\it sp\_minus\_one  $\mapsto$ \it pc\_high \rm ) \rm \}\rm )$\para$

@@

\textbf{RESET}  - The reset is other external action that just reset the
registers related to the interruptions.

@@

\section{Extern  actions and input and output ports}

The Z80 has an extensive set of input and output instruction and 256 ports for
devices. This model can transfer data blocks and between the I/O devices and any
of the internal registers or memory address.

The $\mathit{IN\_r(C)}$ \footnote{The tools B don't allow use parenteses in identifiers, then the characteres
\{``('',``)''\} are replaced respectively by \{``9'',``0''\} in the real specification.} is represented by the follow B
operation. This instruction is from I/O group and update the flag registers.

@@

\subsubsection{Instruction set} 
% Talk in general and put a example with midlle
% complex

@@

%  Para a modelagen dessa plataforma, essa modelagen contém linhas e $np $
%  provas não obvias. Uma forma de acelerar o processo de prova, foi paralelizar
%  as provas. Isto foi feito em 4 IMacs e consumiu $ tp $ tempo para verificar
%  automaticamente $np - p $. As demais foi necessário uma certa interação com
%  o provador para sua verificação.
%  
%  
%  Algumas informações do manual do Z80 fornecido pela Zilog são imprecisas,
%  o que compremete um pouco a qualidade do modelo, entretanto alguns pontos
%  foram consultados no documento (Undocument Z80 documented-(Sean Young)) e
%  dúvidas foram sanadas com o uso do simulador(Z80 IDE Simulator). 
%  
%  Para oferecer uma maior garantia do modelo com a implementação do processador
%  algumas atividades foram realizadas. Primeiramente, tentantamos animar e gerar
%  código respectivamente nas ferramentas ProB e AtelierB. Entratanto, o modelo
%  cotém algumas construções complexas que ainda não são bem suportada por elas.
%  Então, foi realizada uma apresentação sobre a descrição do modelo e foi
%  realizada um nova revisão com 3 pessoas.

\paragraph{Notes and Comments.}

% ---- Bibliography ----
%
\begin{thebibliography}{5}

@@

% Outros do artigo de bartira

@@

\bibitem {Dantas}
 Dantas, B; D\'{e}harbe, D; Galv\~{a}o, S; ET AL. Proposta e avaliação de uma abordagem de desenvolvimento de
 software fidedigno por construção com o método B. In: SEMISH, Belém, 2008. SBC.

>>>>

\usepackage{graphicx,array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

@@

\begin{abstract}
This paper describes an approach to model the microcontroller platforms. More specifically, it shows details about
the Z80 model. The model is created using the B method; it applies math and logic concepts to describe the
characteristics from platforms. Therefore, this model can be used in platforms projects to document, build
simulators, verify properties about the model and verify the software in assembly level. The verification in
assembly is the feature more important, because this allow the developing software faithful to the algorithm
language. Finally, this paper show some relevant techniques used in models and to describe several details.
\end{abstract}
%
\section{Introduction}
% Explica o contexto geral do artigo
%  Porque modelar os microcontroladores. \footnote{The interested reader in details from full specification or in
% other study case is invited to visit the repository from authors at \url{http://b2asm.googlecode.com}.
%    Porque podemos verificar várias propriedades em tempo de projeto, documentar, simulacao através das ferramentas de suporte do método B...
%    Porque permite verificar software em nível de assemblagem.[citar trabalho de BHDL]
%    Porque utilizamos o método B puro  - [citar proposta do ALEMAO do ProB]
%  Contextualizacao com grande desafio da verificao formal em nivel de assemblagem[citar grande challenge] 
%  Conclusão da introducao.
%  Organizacao do paper
%  Basear-se no Relatorio técnico e no ERMAC
%

@@

\subsubsection{Desing components B }
[half page]
[Explicar como as coias são modeladas]
% Use the tech report

@@

The main module includes a instance of memory module and accesses the definitions from: basic data types modules,
\textit{ALU} and \textit{Power2} \footnote{The \textit{Power2} module has the basics definitions to help the
theorems prover about calculus of power.}.

@@

There are many functions defined in the module \textit{ALU}. In general, these functions take basic definitions to build
new specific functions. The function $\mathit{half8UCHAR}$ is used to get the half part of $\mathit{UCHAR}$ value.
It is importante to know the half carry and it is used in the $\mathit{add8UCHAR}$.

@@

Every logic functions are defined in the \textit{BYTE} and \textit{BV16} module are included in the \textit{ALU}
module. Then, it was not necessary rewrite these functions.

\subsection{Modelling the instructions}[I need review the english]

The main module (\textit{Z80}) has two types of operations, a type represent the external actions and the other
represent the instructions of microcontrollers. The external actions are shown in the section
~\ref{sec:externalactions} and the other type represents the instructions of microcontroller. A simple example from
instruction is a $\mathit{LD\_(nn)\_A}$ shown below. Many times to model a instruction is necessary to use the pre-defined functions,
this facilitates the construction of model. This instruction use the $\mathit{updateAddressMem}$ function from
\textit{Memory} module and it receives a address memory and its new memory value. After, it increments the program
counter ($\mathit{pc}$) and update the refresh register ($\mathit{r\_}$).

\hspace*{0.00in}\bf LD\_9nn0\_A \rm ( \it nn \rm ) \rm =

\hspace*{0.20in}\bf PRE \it nn $\in$ \it USHORT\hspace*{0.15in} $\land$ \hspace*{0.10in}\it nn\hspace*{0.10in} $\in$  \it DATA\_R\_ADR

\hspace*{0.20in}\bf updateAddressMem \rm ( \it ushort\_bv16 \rm ( \it nn \rm ) \rm , \it rgs8 \rm ( \it a0 \rm )
\rm )  $\para$

\hspace*{0.20in}\it pc \rm := \it instruction\_next \rm ( \it pc \rm )  $\para$  \it r\_ \rm := \it update\_refresh\_reg\rm (\it r\_\rm )

\hspace*{0.00in}\bf END\rm ;

The other instructions have a similar structure.

\subsection{Modelling the external action}
\label{sec:externalactions}

The external actions change the state of microcontroller and they aren´t instructions, for example, the refreshing
of I/O port and the interruptions request. The extern action are also model by operations and they are prefixed
with ``$ext\_$'' and after the name of action. There are just four extern actions: $ext\_update\_io\_ports$,
$ext\_NMI$ and $ext\_INT$, $ext\_Reset$. The $ext\_update\_io\_ports$ just change the state of I/O port, see.

\hspace*{0.20in}\bf ext\_update\_io\_ports\rm (\it address\rm ,\it value\rm )\rm =

\hspace*{0.20in}\bf PRE \it address  $\in$  \it UCHAR  $\land$ \hspace*{0.10in}\it value  $\in$  \it SCHAR \bf THEN

\hspace*{0.40in}\it io\_ports \rm ( \it uchar\_byte \rm ( \it address \rm ) \rm ) \rm := \it schar\_byte \rm ( \it
value \rm )

\hspace*{0.20in}\bf END\rm

The others are related to interrupts. The interrupts allows that the devices suspend a routine from CPU and start
another service routine. This service routine can exchange data or signals between CPU and external devices. When a
routine is finished, then the CPU come back to the last routine that was interrupted.

@@

\begin{sloppypar}
\bf updateStack\rm (\rm \{ \rm (\it sp\_minus\_two  $\mapsto$  \it pc\_low\rm )\rm ,\rm (\it sp\_minus\_one  $\mapsto$ \it pc\_high \rm ) \rm \}\rm )$\para$

@@

\textbf{RESET}  - This just reset the registers related to the interruptions.

@@

\section{Input and output ports}

The Z80 has an extensive set of input and output (I/O) instruction and 256 ports for
devices. This model can transfer data blocks and between the I/O devices and any
of the internal registers or memory address.

The $\mathit{IN\_r(C)}$ \footnote{The tools B don't allow use parenteses in identifiers, then the characteres
\{``('',``)''\} are replaced respectively by \{``9'',``0''\} in the real specification.} instruction is represented
by the follow B operation. It is from I/O group, then it reveives a identifier of register
$\mathit{r}$ and, in this place, it stores the value of $\mathit{C}$ port address. Besides, it increments the
program counter and updates the flag registers.

@@

\section{Proofs}

[O que as provas garantem?]
As obrigações de provas geradas pela ferramentas garante a verficação dos tipos de dados e que as
expressoes são bem formadas \footnote{An expression is called "well-defined" (or unambiguous) if its definition
assigns it a unique interpretation or value.}. Deste modo, as propriedades citadas anteriomente também são
verificadas. As propriedades citadas anteriormente oferecem garantias que

[Enfatizar a vantagem]
The traditional don´t allow verify this things and the properties is flexible. The flexbility allow add
restrictions about tempo de execuçao and until specifics restrictions about the application of microcontroller.

This flexibility of create propertis allow create conditions to verify if the software to finish in cert instant.

[Estatisticas]
Types Library

Hardware Library

[Groups]
\begin{center}

\begin{tabular}{llll}
\textbf{Group} &  \textbf{Proofs No Obvios} & \textbf{Proofs WD} & × \\ 
Input and Output & 54 & 296  \\ 
Logic and Ari & 134 & 413  \\ 
Manipulation of Bits & 160 & 742  \\ 
Extern actions & 51 & 84  \\
General & 140 & 615  \\
Header & 68 & 169  \\
 & \textbf{Subtotal} 607 & \textbf{Subtotal} 2319  &  \textbf{Total} 2926 \\
\end{tabular}
\end{center}

[Conclusao]
A primeira vez que realiza as provas é um pouco trabalhoso, porque as definições de tipos e de hardware não
estavam completas. Então, um trabalho maior foi demandado, por outro lado, foi nece

The first time
Paralelismo vantagem

@@

[]

%  Para a modelagen dessa plataforma, essa modelagen contém linhas e $np $
%  provas não obvias. Uma forma de acelerar o processo de prova, foi paralelizar
%  as provas. Isto foi feito em 4 IMacs e consumiu $ tp $ tempo para verificar
%  automaticamente $np - p $. As demais foi necessário uma certa interação com
%  o provador para sua verificação.

%  Para oferecer uma maior garantia do modelo com a implementação do processador
%  algumas atividades foram realizadas. Primeiramente, tentantamos animar e gerar
%  código respectivamente nas ferramentas ProB e AtelierB. Entratanto, o modelo
%  cotém algumas construções complexas que ainda não são bem suportada por elas.
%  Então, foi realizada uma apresentação sobre a descrição do modelo e foi
%  realizada um nova revisão com 3 pessoas.

\section{Acknowledges}

This work received supports from CNPq and ANP (National Agency of Oil). Besides, the research Bruno Gomes help in
the revision.

\begin{thebibliography}{5}

@@

@@

\bibitem {HOARE}
HOARE, C. A. R. The verifying compiler, a grand challenge for computing research. In: VMCAI, p. 78-78, 2005.

\bibitem{Medeiros}
 Medeiros Jr, Val\'{e}rio; Galv\~{a}o, Stephenson; D\'{e}harbe, David; Modelagem de Microcontroladores em B. In: ERMAC, Natal, 2008. SBMAC.

\bibitem {WORDSWORTH}
WORDSWORTH, J. B. Software Engineering with B. Addison Wesley, Boston, 1996.
